---
title: "Sankey Diagram for FTU Explorer Paper"
author: "Andreas Bueckle"
date: "2024-09-12"
output: html_document
---

## Load libraries

```{r load libraries}
library(tidyverse)
library(networkD3) #for Sankey
library(RColorBrewer) # for plots
```

## Load data

We need the `sankey.csv` file to then preprocess it into nodes and links.

```{r load data}
sankey = read_csv("../output/sankey_for_FTU.csv", locale=locale(decimal_mark = ",", grouping_mark = "."))
```


```{r}
sankey
```

## Create bins for donor age and BMI

### Define functions 
```{r echo=FALSE}
add_bin = function(original_column, bins, labels){
  # Create new column with bins
  result <- cut(as.numeric(original_column), breaks = bins, labels = labels, include.lowest = TRUE)
  
  return(as.character(result))
}
```

### Handle CxG age

### Load CxG developmental stage (not in hra-pop graph yet)
```{r echo=FALSE}

# load sheet with CxG age values
cxg_age = read_csv("data/cxg-development-stage.csv")

# add column headers
names(cxg_age) <- c('unique_dataset_id','donor_developmental_stage_cxg')
```

### Extract numbers and assign as age
```{r echo=FALSE}

# extract age using {stringr}
cxg_age= cxg_age %>% 
  mutate(
    contains_number = str_detect(donor_developmental_stage_cxg, "\\d"),
     extracted_age_number = as.numeric(str_extract(donor_developmental_stage_cxg, "\\d+"))
         ) 

# join main tibble with cxG-age tibble with extracted age numbers, replace donor_age if new age value provided
sankey <- sankey %>%
  left_join(cxg_age, by = "unique_dataset_id") %>% 
  mutate(donor_age = if_else(is.na(extracted_age_number) | contains_number == FALSE, donor_age, extracted_age_number)) %>%
  select(-extracted_age_number, -contains_number)
```

### Add bins
```{r message=FALSE, echo=FALSE}

# create new tibble
sankey_with_bins= sankey 

# Create bins and labels
# age bins see WHO at https://apps.who.int/gho/data/node.searo-metadata.AGEGROUP
bins_age = c(0,1,5,10,15,20,25,30,35,40,45,50,55,60,65,70,75,80,85,90,95,100)  # Define bin edges
labels_age = c("<1 Year", "1-4 Years", "5-9 Years", "10-14 Years", "15-19 years", "20-24 Years", "25-29 Years", "30-34 Years", "35-39 Years", "40-44 Years", "45-49 Years", "50-54 Years", "55-59 Years", "60-64 Years", "65-69 Years", "70-74 Years", "75-79 Years", "80-84 Years", "85-89 Years", "90-94 Years", "95-99 Years") 

# BMI bins
bins_bmi = c(0,18.5, 24.9, 29.9, 30.0)
labels_bmi = c("underweight", "healthy", "overweight", "obese")

sankey_with_bins$donor_age_binned = add_bin(sankey_with_bins
                                          $donor_age, bins_age, labels_age)
sankey_with_bins$donor_bmi_binned = add_bin(sankey_with_bins
                                          $donor_bmi, bins_bmi, labels_bmi)

```

## Create subset and replace NAs, unify unknown values, adjust portal values
```{r echo=FALSE}

organ_not_supported_text = "Organ Not Supported"

subset_sankey = sankey_with_bins %>% 
  select(portal, donor_sex, organ_name, dataset_id, unique_dataset_id, FTU, cell_type_annotation_tool, donor_race, donor_bmi_binned, donor_age_binned, is_rui_registered, is_atlas_dataset) %>% 
  
  # replace NAs
  replace_na(list(donor_sex = "Unknown Sex")) %>% 
  replace_na(list(donor_race = "Unknown Race")) %>% 
  replace_na(list(organ_name = organ_not_supported_text)) %>% 
  replace_na(list(cell_type_annotation_tool = "No Cell Summary")) %>%
  replace_na(list(dataset_id = "No Known Dataset ID")) %>% 
  replace_na(list(unique_dataset_id = "No Known Unique Dataset ID")) %>% 
  
  # replace NAs for binned variables
  replace_na(list(donor_bmi_binned = "Unknown BMI")) %>%
  replace_na(list(donor_age_binned = "Unknown Age")) %>%
  
  # fix value for is_rui_registered
  
  mutate(
  # make organs lower case
  organ_name = str_to_sentence(organ_name),
  
  # make sex lower case
  donor_sex = str_to_sentence(donor_sex)
  ) %>%
  
  mutate(
    # unify left and right kidney
    organ_name = ifelse(organ_name == "left kidney" | organ_name == "right kidney"| organ_name == "Left kidney"| organ_name == "Right kidney", "Kidney", organ_name))
  
  # unify unknown values in race and sex
  subset_sankey$donor_race[subset_sankey$donor_race == tolower("Unknown")] <- "Unknown Race"
  subset_sankey$donor_race[subset_sankey$donor_race == "na"] <- "Unknown Race"
  subset_sankey$donor_race[subset_sankey$donor_race == "n/a"] <- "Unknown Race"
  subset_sankey$donor_sex[subset_sankey$donor_sex == "Unknown"] <- "Unknown Sex"
  
  # unify FTU values
  subset_sankey$FTU[is.na(subset_sankey$FTU)] <- "Unknown"
  
  # fix portal names
  subset_sankey$portal[subset_sankey$portal == "HCA"] <- "CZ CELLxGENE"
  subset_sankey$portal[subset_sankey$portal == "NHLBI/LungMap"] <- "LungMap"
  subset_sankey$portal[subset_sankey$portal == "CxG"] <- "CZ CELLxGENE"
  subset_sankey$portal[grepl("EBI", subset_sankey$portal)] <- "EBI"
  
  # replace portal==HRA with HRA-OMAP
  subset_sankey$portal[subset_sankey$portal == "HRA"] <- "HRA-OMAP"
  
  # turn LGL into meaningful CHAR for rui and atlas
  subset_sankey$is_rui_registered[subset_sankey$is_rui_registered == TRUE] <- "RUI-registered"
  subset_sankey$is_rui_registered[subset_sankey$is_rui_registered == "True"] <- "RUI-registered"
  subset_sankey$is_rui_registered[subset_sankey$is_rui_registered == FALSE] <- "Not RUI-registered"
  subset_sankey$is_rui_registered[subset_sankey$is_rui_registered == "no"] <- "Not RUI-registered"
  
  subset_sankey$is_atlas_dataset[subset_sankey$is_atlas_dataset == TRUE] <- "Atlas Dataset"
  subset_sankey$is_atlas_dataset[subset_sankey$is_atlas_dataset == FALSE] <- "Not Atlas Dataset"
  subset_sankey$is_atlas_dataset[is.na(subset_sankey$is_atlas_dataset)] <- "Not Atlas Dataset"
    
```

##Remove duplictate rows
```{r echo=FALSE}
 
# replace tool with sc-transcriptomics, no cell summary, sc-proteomics
tool_replacement = "sc_transcriptomics with Cell Summary"

subset_sankey$cell_type_annotation_tool[subset_sankey$cell_type_annotation_tool == "azimuth"] <- tool_replacement
subset_sankey$cell_type_annotation_tool[subset_sankey$cell_type_annotation_tool == "celltypist"] <- tool_replacement
subset_sankey$cell_type_annotation_tool[subset_sankey$cell_type_annotation_tool == "popv"] <- tool_replacement
subset_sankey$cell_type_annotation_tool[subset_sankey$cell_type_annotation_tool == "n/a"] <- tool_replacement

duplicates <- subset_sankey[duplicated(subset_sankey),]
duplicates

subset_sankey <- distinct(subset_sankey)
print(subset_sankey)

```

## Temporary fix for SenNet atlas datasets
``` {r echo=FALSE}

subset_sankey$cell_type_annotation_tool[subset_sankey$portal == "SenNet" & subset_sankey$is_atlas_dataset == "Atlas Dataset"] <- tool_replacement
```

## Map race/ethnicity values to race values
```{r echo=FALSE}
# export donor_race as CSV
donor_race = as_tibble(subset_sankey$donor_race %>% unique())

# set header
donor_race = donor_race %>% rename(donor_race = value)

# export
write_csv(donor_race, "output/donor_race.csv")

# read manually mapped CSV file back in
race_mapped = read_csv("data/donor_race_mapped.csv")

# replace race values in Sankey with mapped values
subset_sankey <- subset_sankey %>%
  left_join(race_mapped, by = "donor_race")

# replace donor_race with mapped values
subset_sankey$donor_race = subset_sankey$mapped_donor_race
```

## OPTIONAL: Filter for only atlas data
``` {r OnlyAtlas, echo=FALSE} 
# if(only_atlas) {subset_sankey = subset_sankey %>% filter(is_atlas_dataset == "Atlas Dataset")}

```

## Create nodes
```{r echo=FALSE}
p = subset_sankey %>% 
  group_by(portal) %>% summarize()

d = subset_sankey %>% 
  group_by(donor_sex) %>% summarize()

a = subset_sankey %>% 
  group_by(donor_age_binned) %>% summarize()

b = subset_sankey %>%
  group_by(donor_bmi_binned) %>% summarize()

r = subset_sankey %>% 
  group_by(donor_race) %>% summarize()

o = subset_sankey %>% 
  group_by(organ_name) %>% summarize()

f = subset_sankey %>% 
  group_by(FTU) %>% summarize()

c = subset_sankey %>% 
  group_by(cell_type_annotation_tool) %>% summarize()

rui = subset_sankey %>% 
  group_by(is_rui_registered) %>% summarize()

atlas = subset_sankey %>% 
  group_by(is_atlas_dataset) %>% summarize()

unique_name=list()
unique_name = unlist(append(unique_name, c(p, d, a, b, r, o, f, rui)))
unique_name = list(unique_name)

nodes = as.data.frame(tibble(name = character()))
```

## Create links
 Use this order: 
  * sources
  * donor sex
  * donor age
  * donor BMI
  * donor ethnicity
  * organ
  * cell type annotation tool used
  * RUI registered or not
  * ‘atlas’ dataset or not 
```{r echo=FALSE}

for(u in unique_name){
  nodes = nodes %>% 
    add_row(name=u)
}

nodes$index <- 1:nrow (nodes) 
nodes

nodes$index = nodes$index-1
nodes

portal_sex = subset_sankey %>% 
  group_by(portal, donor_sex) %>% 
  summarize(count=n()) %>% 
  rename(
    source = portal,
    target = donor_sex,
    value=count
  )

sex_age = subset_sankey %>% 
  group_by(donor_sex, donor_age_binned) %>% 
  summarize(count=n()) %>% 
  rename(
    source = donor_sex,
    target = donor_age_binned,
    value=count
  )

age_bmi = subset_sankey %>% 
  group_by(donor_age_binned, donor_bmi_binned) %>% 
  summarize(count=n()) %>% 
  rename(
    source = donor_age_binned,
    target = donor_bmi_binned,
    value=count
  )

bmi_race = subset_sankey %>% 
  group_by(donor_bmi_binned, donor_race) %>% 
  summarize(count=n()) %>% 
  rename(
    source = donor_bmi_binned,
    target = donor_race,
    value=count
  )

race_organ = subset_sankey %>%
  group_by(donor_race, organ_name) %>%
  summarize(count=n()) %>%
  rename(
    source = donor_race,
    target = organ_name,
    value=count
  )

organ_ftu = subset_sankey %>% 
  group_by(organ_name, FTU) %>% 
  summarize(count=n()) %>% 
  rename(
    source = organ_name,
    target = FTU,
    value=count
  )

ftu_rui = subset_sankey %>% 
  group_by(FTU, is_rui_registered) %>% 
  summarize(count=n()) %>% 
  rename(
    source = FTU,
    target = is_rui_registered,
    value=count
  )




prep_links = as.data.frame(bind_rows(portal_sex, sex_age, age_bmi, bmi_race, race_organ, organ_ftu, ftu_rui))
prep_links 

links = prep_links 
```


## Rename node and link tables
```{r echo=FALSE}
names(nodes)[1] = "source"
prep_links = left_join(prep_links, nodes,by="source")

names(nodes)[1] = "target"
prep_links = left_join(prep_links, nodes,by="target")
prep_links

prep_links = prep_links[,c(4,5,3)]
names(prep_links)[1:2] = c("source", "target")
names(nodes)[1] = "name"
```

## Draw the Sankey diagram

```{r echo=FALSE}
p <- sankeyNetwork(Links = prep_links, Nodes = nodes, Source = "source",
                   Target = "target", Value = "value", NodeID = "name", 
                   units = "occurrences", fontSize = 15, nodeWidth = 20, width=2500, height = 900)

p
```

## Export the Sankey diagram as HTML

```{r echo=FALSE}

saveNetwork(p, "../docs/ftu_sankey_atlas.html")

```